SDGs

### 2023-02-28 inequality-SDGs.Rmd
updating pca visualizations

p <- dat |>
    filter(indicator_name %in% maike_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    theme_light(base_size = 6)

# ggsave(
#     plot = p, filename = "WB_environment_SDGs.png", path = "figures/", device = "png",
#     width = 6, height = 4, dpi = 400, bg = "white"
# )

@@

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    theme_light(base_size = 6)

@@

Wait until Maike and Tong revise the selection of variables. Take the analysis from there. Alternatively, select a couple of topics as example, just to explore how to deal with missing values and what is doable. Once one take missing values into account, the number of variables and countries get reduced quickly.

@@

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ungroup() |> group_by(indicator_name) |> 
    summarize(mean_missing = mean(missing)) |> 
    filter(mean_missing < 0.3) |> 
    ggplot(aes(mean_missing, indicator_name)) +
    geom_point() +
    #scale_fill_viridis_c() +
    theme_light(base_size = 6)

@@

q <- dat |>
    filter(indicator_name %in%  vars) |>
    select(-x35) |>
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |>
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |>
    mutate(nas = is.na(value)) |>
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    group_by(country_code, indicator_name) |>
    summarise(missing = sum(nas) / diff(range(year))) |>
    #filter(country_code %in% complete_countries) |> # this line only if next cell already run
    # ungroup() |> group_by(country_code) |>
    # summarize(mean_missing = mean(missing)) |>
    #filter(mean_missing < 0.3) |>
    ggplot(aes(country_code, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    theme_light(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90))

q

@@

@@

```{r warnings=FALSE, message=FALSE}
dat |>
    filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) |> 
    GGally::ggpairs(
        columns = 4:12 ,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 6)
```
Interpolate missing values

@@

sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 10) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    labs(title = "SDG10: Inequality indicators") +
    theme_light(base_size = 8)

# sdg |> 
#     filter(series_description == "Proportion of voting rights of developing countries in international organizations, by organization (%)")

@@

```{r}
sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 14) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    labs(title = "SDG14: Life under water") +
    theme_light(base_size = 8)
```

```{r}
sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    group_by(indicator, indicator_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |>
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator, indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    mutate(indicator_name = str_trunc(indicator_name, width = 30, side = "right")) |>
    ggplot(aes(year, indicator)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    labs(title = "SDG15: Life on land") +
    theme_light(base_size = 8)

@@

@@

```{r}
p2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(goal %in% c(10, 14, 15)) |> 
    group_by(series_code, year, iso3) |>
    # this gets rid of multiple values per year
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    ungroup() |> group_by(series_code, iso3) |> 
   # group_by(indicator, indicator_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year)) ) |>  ## dividing by number of years (per country)
    #pull(missing) |> range() # should be between 0-1
    # summarize(mean_missing = mean(missing)) |> 
    filter(missing < 0.3) |> 
    ungroup() |> 
    pivot_wider(names_from = series_code, values_from = missing) |> 
    pivot_longer(cols = AG_LND_FRSTCERT:last_col(), names_to = "series_code",
                 values_to = "missing") |> 
    group_by(series_code) |> 
    #filter(!any(is.na(missing))) |> 
    # ggplot(aes(iso3, series_code)) +
    # geom_tile(aes(fill = missing)) +
    # theme_light()
    summarize(missing_countries = sum(is.na(missing))) |> 
    filter(missing_countries < 45) |> 
    ggplot(aes(missing_countries, series_code))+
    geom_point()

p2 + theme_light(base_size = 12)

vars2 <- p2$data$series_code
```

```{r}
q2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(series_code %in% vars2) |> 
    ## this line is to get rid of multiple values per year
    group_by(series_code, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> #group_by(indicator_name, year, iso3) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(series_code, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year))) |>  ## dividing by number of years (per country)
    # summarize(mean_missing = mean(missing)) |> 
    ungroup() |> 
    filter(missing < 0.3) |> 
    ## add missing values again for combination of indicator country:
    ungroup() |> group_by(series_code) |>
    pivot_wider(names_from = iso3, values_from = missing) |> 
    pivot_longer(cols = AFG:last_col(), names_to = "iso3", values_to = "missing") |>
    ## completely empty time series
    #summarize(absent_countries = sum(is.na(missing))) |> 
    ungroup() |> group_by(iso3 ) |> 
    #filter(sum(is.na(missing)) < 45) |> # remove absent_countries > 45
    filter(!any(is.na(missing))) |> 
    ggplot(aes(iso3, series_code)) +
    #geom_point()
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c() +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90))

@@

complete_countries_sdg <- q2$data$iso3 |> unique()
```

@@

```{r}
df_un_reduced <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries_sdg) |> # use only countries
    filter(series_code %in% vars2) |> 
    group_by(series_code, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    group_by(iso3, series_code) |> 
    mutate(value = na_interpolation(value, option = "spline"))

@@

@@

```{r}
rm(df_sdg, df_un, df_wb, maike_vars, other_vars)
# key files
# sdg raw data from UN SDG
# dat raw data from WorldBank
```

@@

@@

```{r}
biplot(pca1)
```

```{r}
# summary(pca1)
```

```{r}
pca2 <- df_un_reduced |> 
    unite(col = "var", series_code, year, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    ungroup() |> 
    select(-iso3) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    prcomp(center = TRUE, scale. = TRUE)
```

>>>>

p <- dat |>
    filter(indicator_name %in% maike_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(20, "mm"))) + 
    labs(x = "Year", y = "Indicator name") +
    theme_light(base_size = 6)
# 
# ggsave(
#     plot = p, filename = "WB_environment_SDGs.png", path = "figures/", device = "png",
#     width = 6, height = 4, dpi = 400, bg = "white"
# )

@@

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(20, "mm"))) + 
    labs(x = "Year", y = "Indicator name")
    theme_light(base_size = 6)

@@

**Wait until Maike and Tong revise the selection of variables.** Take the analysis from there. Alternatively, select a couple of topics as example, just to explore how to deal with missing values and what is doable. Once one take missing values into account, the number of variables and countries get reduced quickly.

@@

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ungroup() |> group_by(indicator_name) |> 
    summarize(mean_missing = mean(missing)) |> 
    filter(mean_missing < 0.3) |> 
    ggplot(aes(mean_missing, indicator_name)) +
    geom_point() +
    #scale_fill_viridis_c(
    #    guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(20, "mm"))) + 
    labs(x = "Mean proportion of missing values across countries", y = "Indicator name", tag = "A")+
    theme_light(base_size = 6)

#p + theme_light(base_size = 8)

@@

q <- dat |>
    filter(indicator_name %in%  vars) |>
    select(-x35) |>
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |>
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |>
    mutate(nas = is.na(value)) |>
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    group_by(country_code, indicator_name) |>
    summarise(missing = sum(nas) / diff(range(year))) |>
    #filter(country_code %in% complete_countries) |> # this line only if next cell already run
    # ungroup() |> group_by(country_code) |>
    # summarize(mean_missing = mean(missing)) |>
    #filter(mean_missing < 0.3) |>
    ggplot(aes(country_code, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        "Proportion of missing values",
        guide = guide_colorbar(barwidth = unit(2,"cm"), barheight = unit(2,"mm"))) + 
    labs(x = "Country code", y = "Indicator name", tag = "B") +
    theme_light(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90, size = 3), 
          legend.position = "bottom")

p/q
# ggsave(
#     plot = p/q, filename = "WB_selected_vars.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )

@@

**J230226:** If I add Gini for inequality from the inequality database, our available countries will drop down to 50+, which is what we did in the trilemma paper. It is risky talking about inequality wihtout Gini because the variables selected are not necessarily interpretable as inequality, but let's see what the group supports.

@@

```{r warnings=FALSE, message=FALSE}
crg <- dat |>
    filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) |> 
    GGally::ggpairs(
        columns = 4:12 ,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 4)

crg

# ggsave(
#     plot = crg, filename = "WB_correlogram.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )
```
Interpolate missing values

@@

a <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 10) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) + 
    labs(title = "SDG10: Inequality indicators", 
         x = "Year", y = "Indicator name", tag = "A") +
    theme_light(base_size = 6)
a
# sdg |> 
#     filter(series_description == "Proportion of voting rights of developing countries in international organizations, by organization (%)")

@@

```{r}
b <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 14) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
     scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG14: Life under water", x = "Year", y = "Indicator name", tag = "B") +
    theme_light(base_size = 6)
```

```{r}
c <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    group_by(indicator, indicator_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |>
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator, indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    mutate(indicator_name = str_trunc(indicator_name, width = 100, side = "right")) |>
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c("Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG15: Life on land", x = "Year", y = "Indicator name", tag = "C") +
    theme_light(base_size = 6)
c

@@

# ggsave(
#     plot = a/b/c, filename = "UN_selected_vars.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )

@@

```{r}
p2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(goal %in% c(10, 14, 15)) |> 
    group_by(series_name, year, iso3) |>
    # this gets rid of multiple values per year
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    ungroup() |> group_by(series_name, iso3) |> 
   # group_by(indicator, indicator_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year)) ) |>  ## dividing by number of years (per country)
    #pull(missing) |> range() # should be between 0-1
    # summarize(mean_missing = mean(missing)) |> 
    filter(missing < 0.3) |> 
    ungroup() |> 
    pivot_wider(names_from = series_name, values_from = missing) |> 
    pivot_longer(cols = `Average proportion of Freshwater Key Biodiversity Areas (KBAs) covered by protected areas (%)`:last_col(), names_to = "series_name",
                 values_to = "missing") |> 
    group_by(series_name) |> 
    #filter(!any(is.na(missing))) |> 
    # ggplot(aes(iso3, series_name)) +
    # geom_tile(aes(fill = missing)) +
    # theme_light()
    summarize(missing_countries = sum(is.na(missing))) |> 
    filter(missing_countries < 45) |> 
    ggplot(aes(missing_countries, series_name))+
    geom_point() +
    labs(tag = "A", y = "Series name", x = "Proportion of missing countries") +
    theme_light(base_size = 6)
p2
vars2 <- p2$data$series_name
```

```{r}
q2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(series_name %in% vars2) |> 
    ## this line is to get rid of multiple values per year
    group_by(series_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> #group_by(indicator_name, year, iso3) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(series_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year))) |>  ## dividing by number of years (per country)
    # summarize(mean_missing = mean(missing)) |> 
    ungroup() |> 
    filter(missing < 0.3) |> 
    ## add missing values again for combination of indicator country:
    ungroup() |> group_by(series_name) |>
    pivot_wider(names_from = iso3, values_from = missing) |> 
    pivot_longer(cols = AFG:last_col(), names_to = "iso3", values_to = "missing") |>
    ## completely empty time series
    #summarize(absent_countries = sum(is.na(missing))) |> 
    ungroup() |> group_by(iso3 ) |> 
    #filter(sum(is.na(missing)) < 45) |> # remove absent_countries > 45
    filter(!any(is.na(missing))) |> 
    ggplot(aes(iso3, series_name)) +
    #geom_point()
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c("Missing",
        guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(2,"cm"))) +
    labs(tag = "B", y = "Series name", x = "Countries") +
    theme_light(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90))

@@

complete_countries_sdg <- q2$data$iso3 |> unique()

# ggsave(
#     plot = p2/q2, filename = "UN_reduced_vars.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )

@@

```{r}
df_un_reduced <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries_sdg) |> # use only countries
    filter(series_name %in% vars2) |> 
    group_by(series_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    group_by(iso3, series_name) |> 
    mutate(value = na_interpolation(value, option = "spline"))

@@

```{r}
crg <- df_un_reduced |> 
    mutate(series_name = str_replace(series_name, "Biodiversity", "Biodiversity\n"),
           series_name = str_replace(series_name, "verified", "verified\n")) |> 
    pivot_wider(values_from = value, names_from = series_name) |> 
    GGally::ggpairs(
        columns = 3:8,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 4)

crg

# ggsave(
#     plot = crg, filename = "UN_correlogram.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )
```

@@

```{r}
rm(df_sdg, df_un, df_wb, maike_vars, other_vars)
# key files
# sdg raw data from UN SDG
# dat raw data from WorldBank

@@

Following Ingo's advise, here I run a PCA with each variable_year as variable.

@@

```{r}
# some notes on the pca object
pca1$rotation |> dim() # 171 variables*years, 160 PCs because there was 160 countries to start with.
pca1$x |> dim() # The loadings of 160 countries in each of the 160 components.
pca1$center |> length() # 171 variable centers
pca1$scale |> length() # 171 variable scales.
```

```{r}
pca1$rotation |> 
    as_tibble(rownames = "vars") |> 
    separate(vars, sep = "_", into = c("year", "variable")) |> 
    mutate(year = as.numeric(year)) |> 
    pivot_longer(cols = starts_with("PC"), names_to = "PC", values_to = "load") |> 
    mutate(PC = str_remove(PC, "PC"), PC = as.numeric(PC)) |> 
    filter(PC < 4) |> 
    ggplot(aes(year, load)) +
    geom_line(aes(color = variable)) +
    facet_wrap(~PC) +
    theme_light(base_size = 12)

```{r}
biplot(pca1, var.axes = FALSE)
```

```{r}
x <- summary(pca1, loadings = F, cutoff = 0.1)
x <- x$importance |> as_tibble(rownames = "var") |> 
    pivot_longer(cols = starts_with("PC"), names_to = "comp") |> 
    mutate(comp = str_remove(comp, "PC") |> as.numeric()) |> 
    dplyr::filter( comp <= 10 , var != "Standard deviation")

ggplot(data = x |> filter(var == "Proportion of Variance"), aes(comp, value)) +
    geom_col(color = "grey25") +
    # geom_point(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    # geom_line(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    labs(x = "Principal components", y = "Explained variance") +
    scale_x_continuous(n.breaks = 10) +
    scale_y_continuous(labels = scales::label_percent()) +
    theme_light(base_size = 6)

pca1$x |> as_tibble() |> 
    select(PC1:PC2) |> 
    mutate(iso = df_dat$country_code |> unique()) |> 
    ggplot(aes(PC1, PC2)) +
    geom_text(aes(label = iso), size = 2) +
    geom_segment(
        data = pca1$rotation |>
            as_tibble(rownames = "factor") |>
            separate(factor, into = c("year", "factor")),
        aes(xend = PC1, yend = PC2, x= 0, y = 0, color = factor, alpha = year)) +
    coord_equal()

```{r}
pca2 <- df_un_reduced |> 
    unite(col = "var", series_name, year, sep = "_") |> 
    pivot_wider(values_from = value, names_from = var) |> 
    ungroup() |> 
    select(-iso3) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    prcomp(center = TRUE, scale. = TRUE)
```
